---
title: "Getting started with JMnorm"
date: "2023-03-31"
output:
  html_document: default
  pdf_document: default
  word_document: default
---
### JMnorm: a novel approach for Jointly Multi-feature normalization of epigenomic data across cell types and species
***
##### Guanjue Xiang
##### CAMP4 Therapeutics Corp., Cambridge, MA, USA

***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Description: This R Markdown file demonstrates the use of JMnorm to normalize the target signal matrix "TCD8.raw_sigmat.txt" against the reference signal matrix "ref.raw_sigmat.txt". The input matrices, "TCD8.raw_sigmat.txt" and "ref.raw_sigmat.txt", should be formatted as N-by-(M+1) matrices, where N represents the number of cCREs, and M represents the number of chromatin features. The first column of each matrix contains the cCRE IDs. The signal values for each chromatin feature in the cCREs are expected to be non-negative and in linear scale.

***

##### run JMnorm
***
Load Libraries
```{r}
start.time <- Sys.time()
library(readr)
library(pheatmap)
library(dynamicTreeCut)
```

Load JMnorm source code & setup working directory.

!!! Change the file path of "JMnorm.script.R" and working directory path.
```{r}
# Source the JMnorm script 
source('/Users/guanjuexiang/Documents/projects/git/JMnorm/bin/JMnorm_core/JMnorm.script.R')

# Set the working directory
setwd('/Users/guanjuexiang/Documents/projects/git/JMnorm/test_data/')
```

Set parameters
```{r}
# get feature list
feature_list = c("ATAC", "H3K27ac", "H3K27me3", "H3K36me3", "H3K4me1", "H3K4me3", "H3K9me3")

# added 1 to avoid 0 in log transformation
add_sn = 1

# input file names
# The testing signal matrices can be found in this link: https://github.com/guanjue/JMnorm/tree/main/docs
refer_sigmat_file = 'ref.raw_sigmat.txt'
target_sigmat_file = 'TCD8.raw_sigmat.txt'
after_JMnorm_output_target_sigmat_file = 'TCD8.JMnorm_sigmat.txt'
```

Read reference signal matrix
```{r}
# read refe_ave_sigmat
file_ref = as.data.frame(read_table(refer_sigmat_file, col_names = T))
file_ref_sig0 = file_ref[,-1]
```

Normalize signal to noise ratio (SNR) of reference signal matrix
```{r}
# normalize signal to noise ratio (SNR) across features (Optional)
file_ref_sig = normalize_snr(file_ref_sig0)
# if user decided to NOT normalize SNR across different features, then use the following command instead
# file_ref_sig = file_ref_sig0

# log2 transformation
file_ref_sig_log2 = log2(file_ref_sig+add_sn)
```

Determine number of clusters in reference signal matrix
```{r}
# Determine the K value for JMnorm based on reference signal matrix
K = determineK(file_ref_sig_log2, sample_size=20000)
```

Read target signal matrix
```{r}
# read target signal matrix
target_ct = as.data.frame(read_table(target_sigmat_file, col_names = T))

# add noise to avoid many 0s
target_ct_sig0 = target_ct[,-1]+matrix(runif(prod(dim(target_ct)), min = -0.00001, max = 0.0001), nrow = dim(target_ct)[1])

# avoid negative values
target_ct_sig0 = target_ct_sig0-min(target_ct_sig0)
```

Normalize signal to noise ratio (SNR) of target signal matrix
```{r}
# normalize signal to noise ratio (SNR) across features
target_ct_sig = normalize_snr(target_ct_sig0)

# log2 transformation
target_ct_sig_log2 = log2(target_ct_sig+add_sn)
```

Use JMnorm to normalize target signal matrix against reference signal matrix
```{r}
# JMnorm
target_ct_sig_JMnorm = JMnorm(target_ct_sig_log2, file_ref_sig_log2, K)
```

##### Write JMnorm normalized signal matrix
***
```{r}
# add colnames
colnames(file_ref_sig_log2) = feature_list

# correlation between features in reference cell-type
file_ref_sig_log2_cor = cor(file_ref_sig_log2)

# add colnames
colnames(target_ct_sig_JMnorm) = feature_list

# convert JMnorm normalized target signal matrix back to linear scale
target_ct_sig_JMnorm_linear = 2^(target_ct_sig_JMnorm)-add_sn

# write output JMnorm normalized target signal matrix
output_mat = cbind(target_ct[,1], round(target_ct_sig_JMnorm_linear, 3))
colnames(output_mat) = colnames(target_ct)
write.table(output_mat, after_JMnorm_output_target_sigmat_file, quote=F, col.names=T, row.names = F, sep='\t')
```


##### Plot cross-feature correlation matrix heatmap
*** 
Get cross-feature correlation matrix
```{r}
# correlation between features in target cell-type before JMnorm
target_ct_sig_cor = cor(target_ct_sig_log2)

# correlation between features in target cell-type after JMnorm
target_ct_sig_JMnorm_cor = cor(target_ct_sig_JMnorm)

# calculate the R2 between the correlation matrix of target cell-type and reference cell-type
r2_after = r2(as.numeric(target_ct_sig_JMnorm_cor), as.numeric(file_ref_sig_log2_cor))

r2_before = r2(as.numeric(target_ct_sig_cor), as.numeric(file_ref_sig_log2_cor))
```

```{r}
print(paste0('R2 of correlation matrix between target reference before JMnorm: ', r2_before))
print(paste0('R2 of correlation matrix between target reference after JMnorm: ', r2_after))
```


(1) plot the correlation matrix of target cell-type AFTER JMnorm
```{r fig1, fig.width=5., fig.height=6}
par(mfrow=c(1,2))
# plot the correlation matrix of target cell-type after JMnorm
breaksList = seq(-1,1, by = 0.001)
my_colorbar=colorRampPalette(c('blue', 'white', 'red'))(n = length(breaksList))
pheatmap(target_ct_sig_JMnorm_cor, color=my_colorbar, breaks = breaksList, cluster_rows = F, cluster_cols = F, show_rownames = F, show_colnames = T, cex=1.2, main='Target Corr matrix (After JMnorm)')
```

(2) plot the correlation matrix of target cell-type BEFORE JMnorm
```{r fig2, fig.width=5., fig.height=6}
# plot the correlation matrix of target cell-type before JMnorm
colnames(target_ct_sig_cor) = feature_list
pheatmap(target_ct_sig_cor, color=my_colorbar, breaks = breaksList, cluster_rows = F, cluster_cols = F, show_rownames = F, show_colnames = T, cex=1.2, main='Target Corr matrix (Before JMnorm)')
```

(3) plot the correlation matrix of reference
```{r fig3, fig.width=5., fig.height=6}
breaksList = seq(-1,1, by = 0.001)
my_colorbar=colorRampPalette(c('blue', 'white', 'red'))(n = length(breaksList))
# plot the correlation matrix of reference cell-type
pheatmap(file_ref_sig_log2_cor, color=my_colorbar, breaks = breaksList, cluster_rows = F, cluster_cols = F, show_rownames = F, show_colnames = T, cex=1.2, main='Reference Corr matrix')
```

(4) plot scatterplots of correlations between reference and target (Before & After JMnorm)
```{r fig4, fig.width=9, fig.height=5}
par(mfrow=c(1,2))

# plot the correlation matrix of reference cell-type
plot(as.numeric(file_ref_sig_log2_cor), as.numeric(target_ct_sig_JMnorm_cor), xlab='Reference', ylab='Target', xlim=c(-0.5,1), ylim=c(-0.5,1), main='After JMnorm', cex.axis=1.5)
abline(0,1)

plot(as.numeric(file_ref_sig_log2_cor), as.numeric(target_ct_sig_cor), xlab='Reference', ylab='Target', xlim=c(-0.5,1), ylim=c(-0.5,1), main='Before JMnorm', cex.axis=1.5)
abline(0,1)
end.time <- Sys.time()
```


```{r}
print(paste0('Running time on a MacBookPro 2021 M1Pro Chip 16GB REM: ', round((end.time - start.time), 3)))
```


***
##### Reference
*** 
Guanjue Xiang, Yuchun Guo, David Bumcrot. JMnorm: a novel approach for Jointly Multi-feature normalization of epigenomic data across cell types and species (2023)

***








